---
title: "Cyentia Breaches Data"
author: "Joran Elias"
format: html
execute:
  echo: false
  message: false
  warnings: false
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(tidyverse)
  library(skimr)
  library(ggrepel)
  library(patchwork)
  library(quantreg)
  library(broom)
  library(gt)
})

theme_set(theme_bw())

breaches <- readr::read_csv(
  file = "../input/sample_breaches.csv",
  show_col_types = FALSE
) |>
  tidyr::separate(
    col = "cause",
    into = c("internal_external", "cause"),
    sep = " - "
  ) |>
  mutate(breach_year = lubridate::year(breach_date)) |>
  mutate(across(starts_with("naic"), as.character))
```

### Data summary

The `skimr` package provides a convenient, quick overview of the data.

```{r}
#| label: summary
#| echo: true
breaches_skim <- skimr::skim(breaches, -c(id, breach_year))
breaches_skim
```

A few things stand out:

* `affected_count` & `total_amount` are extremely skewed and there are breaches 
that apparently impacted zero records and had zero dollar cost (`r scales::percent_format(accuracy = 0.1)(with(breaches,mean(affected_count == 0)))` of all breaches)
* `sector` & `cause` both are missing in some cases

Additionally, there is a near equal distribution of breaches by `cause` and by `year`:

:::: {.columns}
::: {.column}
```{r cause_dist}
# With the exception of Other, there are ~700 breaches for each
# cause
breaches |>
  summarise(
    n = n(),
    .by = "cause"
  ) |>
  arrange(desc(n))
```
:::

::: {.column}
```{r year_dist}
# There are exactly 10k records in the file and nearly 1k breach records
# per year, which seems suspiciously even
breaches |>
  summarise(
    n = n(),
    .by = "breach_year"
  ) |>
  arrange(breach_year)
```
:::
::::

This makes me wonder if the distribution of the raw number of breaches by `cause` 
and over time are representative samples or if they're an artificially balanced 
subsample from a larger data set. As a result, I suspect that comparisons of simple counts of 
breaches over time or the number breaches between different `cause` will be 
unlikely to show significant differences.

### Breach cost vs affected records by NAIC sector
First we look at potential differences between NAIC sectors with respect to:

* Percent of breaches with nonzero cost/affected records
* Number of breaches with nonzero cost/affected records
* Relationship between breach cost & number of affected records

```{r}
#| label: cost_vs_records
#| fig-width: 12
#| fig-height: 7
# Amt vs Records for breaches with >0 impact
amt_vs_rec <- breaches |>
  filter(total_amount > 0 & affected_count > 0) |>
  summarise(
    n_breaches = n(),
    med_amt = median(total_amount, na.rm = TRUE),
    med_rec = median(affected_count, na.rm = TRUE),
    .by = "sector"
  ) |>
  tidyr::replace_na(replace = list(sector = "Unknown"))

p1 <- ggplot(
  data = amt_vs_rec,
  aes(x = med_rec, y = med_amt, size = n_breaches)
) +
  geom_point(color = "darkblue") +
  geom_text_repel(aes(label = sector, size = NULL)) +
  scale_y_continuous(labels = scales::dollar_format(scale = 0.001, suffix = "K")) +
  ggtitle(
    label = "Breach cost vs afftected records by NAIC sector",
    subtitle = "Only breaches with nonzero cost are included"
  ) +
  labs(
    x = "Median records affected",
    y = "Median Cost",
    size = "# of nonzero cost breaches"
  ) +
  theme(
    legend.direction = "horizontal",
    legend.position = "bottom", plot.title = element_text(vjust = 3.5)
  )

prop_zero_cost <- breaches |>
  summarise(
    prop0 = mean(total_amount == 0),
    .by = "sector"
  ) |>
  tidyr::replace_na(replace = list(sector = "Unknown")) |>
  arrange(prop0) |>
  mutate(sector = fct_inorder(sector))

p2 <- ggplot(data = prop_zero_cost, aes(x = sector, y = prop0)) +
  geom_segment(aes(x = sector, xend = sector, y = 0, yend = prop0), color = "grey") +
  geom_point(color = "darkblue", size = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  ggtitle(label = "% of breaches with\nzero affected records") +
  labs(x = NULL, y = "Percent") +
  coord_flip() +
  theme(panel.grid = element_blank())

p1 + p2 + plot_layout(widths = c(2, 1))
```

The Healthcare, Financial, Administrative, Education & Professional sectors 
appear to be the biggest targets, with the most breaches.

The Agriculture & Transportation sectors appear to have breaches that tend to incur
much higher costs than other sectors, relative to the number of affected records. 

The Other sector tends to have breaches that incur average costs but involve much
higher than normal numbers of records. The breaches in this sector appear to be 
dominated by organizations like:

* Civic & Social Orgs (alumni assoc, ethnic assoc, PTAs, veterans assoc)
* Social Advocacy Orgs 
* Business & Professional Orgs (lawyer/doctor/dentist orgs, trade groups, chambers of commerce)

These are all organizations that likely store large amounts of member information but 
may not see themselves as likely cyberattack targets, or may have extremely 
limited IT security resources at their disposal.

Most of the sectors have roughly similar proportions of breaches with zero
affected records, with the exception of Entertainment, Utilities & Agriculture.

### Cost by cause
Evaluating breach cost is complicated by its skewed nature. There are 
a significant number of breaches with zero cost and the occasional breach with
extremely large costs. We'd like to convey both what a "typical" risk is and also
what the extreme end of the risk may be. To do this, we can show a scatterplot
of the 50th & 95th quantiles by breach cause.

```{r}
#| label: cost_per_breach
cost_per_breach <- breaches |>
  summarise(
    q_mid = quantile(total_amount, 0.5),
    q_hi = quantile(total_amount, 0.95),
    .by = c("cause")
  ) |>
  tidyr::replace_na(replace = list(cause = "Unknown"))

ggplot(data = cost_per_breach, aes(x = q_mid, y = q_hi)) +
  geom_point(color = "darkblue") +
  geom_text_repel(aes(label = cause)) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format(scale = 0.001, suffix = "K")) +
  ggtitle(label = "Median vs Extreme Breach Cost by Cause") +
  labs(
    x = "Median Breach Cost",
    y = "95th Percentile Cost"
  )
```

Breaches caused by former consultant have some of the 
highest median costs per breach, but appear less likely to generate extremely high
cost breaches compared to other causes.

Beaches caused by employees, former employees or trusted third parties have 
somewhat lower median costs, but the risk of an extremely expensive breach from these sources
is considerably higher.

Hacktivist breaches are arguably the worst, as they have relatively high median 
*and* extreme breach costs.

### Trends in breach cost by cause
Again, because the breach cost variable `total_amount` is so heavily skewed, we will 
estimate trends of quantiles, again focusing on an estimate of the "typical"
cost (median) as well as an "extreme" cost (95%-tile).

One option for accomplishing this is to fit a linear quantile regression model 
on the log transformed breach cost: 

$$
log(total\_amount + 1) = \beta_0 + \beta_1 breach\_date + \beta_2 cause + \beta_3(breach\_date*cause)
$$

Quantile regression models estimate the conditional quantiles of the response variable 
instead of the conditional mean, as in a traditional linear model. In this case the 
breach cause specific trend estimate is the sum of the coefficients $\beta_1 + \beta_3$ 
for each `cause`.

```{r}
#| label: quantile-reg
#| cache: true
# Helper function for bootstrapping CIs
cause_coef <- function(model) {
  coef(model) |>
    as.data.frame() |>
    tibble::rownames_to_column(var = "term") |>
    filter(grepl("^breach_date", term)) |>
    set_names(nm = c("term", "0.5", "0.95")) |>
    pivot_longer(cols = -c(term), names_to = "tau", values_to = "estimate") |>
    group_by(tau) |>
    mutate(estimate = estimate + estimate[term == "breach_date"]) |>
    ungroup() |>
    filter(term != "breach_date")
}

# Fill in NAs and log transform total_amount
breaches_reg <- breaches |>
  tidyr::replace_na(replace = list(sector = "Unknown", cause = "Unknown")) |>
  mutate(log_total_amount = log(total_amount + 1))

# Quantile regression model
m <- rq(log_total_amount ~ breach_date + cause + breach_date:cause,
  tau = c(0.5, 0.95),
  data = breaches_reg
)

# Sort of crude bootstrapped CIs for the cause-specific slope
bs_ci <- vector(mode = "list", length = 1000)
for (i in seq_along(bs_ci)) {
  suppressWarnings({
    m_i <- rq(log_total_amount ~ breach_date + cause + breach_date:cause,
      tau = c(0.5, 0.95),
      data = breaches_reg[sample(10000, 10000, replace = TRUE), ]
    )
  })
  bs_ci[[i]] <- cause_coef(m_i)
}
bs_ci <- bind_rows(bs_ci)
ci <- bs_ci |>
  summarise(
    lo = quantile(estimate, 0.025),
    hi = quantile(estimate, 0.975),
    .by = c("term", "tau")
  )

# Fitted values for plotting
m_aug <- augment(m)

# Model
m_tidy <- tidy(m)
m_tidy_cause <- m_tidy |>
  mutate(tau = as.character(tau)) |>
  filter(grepl("breach_date", term)) |>
  group_by(tau) |>
  mutate(estimate = estimate + estimate[term == "breach_date"]) |>
  ungroup() |>
  select(term, tau, estimate) |>
  filter(term != "breach_date")
```

#### Coefficient summary

The resulting trend estimates are summarised in the plot and table below.

:::: {.columns}
::: {.column}
```{r}
#| label: coef-summary-plt
#| fig-height: 13.5
#| fig-width: 8
cause_coefs <- left_join(
  m_tidy_cause |> mutate(tau = as.character(tau)),
  ci,
  by = c("term", "tau")
) |>
  mutate(term = gsub("^breach_date:cause", "", term)) |>
  mutate(across(where(is.numeric), ~ .x * 365.25)) |>
  arrange(desc(term)) |>
  mutate(term = fct_inorder(term))

ggplot(
  data = cause_coefs,
  aes(x = term, y = estimate, color = tau)
) +
  geom_hline(yintercept = 0) +
  geom_pointrange(aes(ymin = lo, ymax = hi),
    size = 0.3,
    linewidth = 0.8,
    position = position_dodge(width = 1)
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = NULL,
    y = NULL,
    color = "Cost Quantile"
  ) +
  coord_flip() +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14)
  )
```
:::

::: {.column}
```{r}
#| label: coef-summary-tbl

cause_coefs <- left_join(
  m_tidy_cause |> mutate(tau = as.character(tau)),
  ci,
  by = c("term", "tau")
)

cause_coefs |>
  mutate(term = gsub("^breach_date:cause", "", term)) |>
  mutate(across(where(is.numeric), ~ .x * 365.25)) |>
  mutate(across(where(is.numeric), ~ scales::percent_format(accuracy = 0.1)(.x))) |>
  mutate(est_lbl = paste(estimate, paste0("(", lo, ",", hi, ")"))) |>
  select(term, tau, est_lbl) |>
  mutate(tau = paste("Quantile:", tau)) |>
  pivot_wider(names_from = tau, values_from = est_lbl) |>
  gt() |>
  tab_options(table.font.size = pct(65), table.width = pct(80)) |>
  tab_header(
    title = "Avg Annual % Change in Breach Cost",
    subtitle = "Estimated change in median & 95%-tile total breach cost."
  ) |>
  cols_label(term = "Cause") |>
  tab_style(locations = cells_title(), style = cell_text(align = "left")) |>
  tab_style(
    locations = cells_body(columns = 2, rows = 3),
    style = cell_text(style = "italic", weight = "bold",color = "darkorange")
  ) |>
  tab_style(
    locations = cells_body(columns = 3, rows = 8),
    style = cell_text(style = "italic", weight = "bold",color = "darkorange")
  ) |>
  tab_style(
    locations = cells_body(columns = 2, rows = 1),
    style = cell_text(style = "italic",color = "darkorange")
  ) |>
  tab_style(
    locations = cells_body(columns = 3, rows = 5:7),
    style = cell_text(style = "italic",color = "darkorange")
  )
```
:::
::::

Only two `cause` specific trends in breach cost appear statistically significant:

* Median breach cost caused by Former Consultants has decreased at an average rate of `-12.3%` per year
* The 95%-tile breach cost with a cause of Other has increased at an average rate of `16.2%` per year

While the remaining trends don't appear statistically significant, several of the 
estimated trends are fairly large. These include large estimated growth rates in the
95%-tile breach cost caused by Nation States, Hacktivists & Organizations.


#### Examples of breach causes with large 95%-tile changes
Finally, we can visualize some of the larger trend estimates, on a log scale.
```{r}
#| label: estimated-quantile-plt
#| fig-height: 8
#| fig-width: 8
m_aug |>
  filter(cause %in% c("Other", "Hacktivist", "Nation State", "Organization", "Former Consultant")) |>
  ggplot(data = _, aes(x = breach_date)) +
  facet_wrap(~cause, ncol = 2) +
  geom_point(aes(y = log_total_amount), alpha = 0.1) +
  geom_line(aes(y = .fitted, color = .tau), linewidth = 1.2) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = "log(Breach Cost)", color = "Quantile") +
  theme(legend.direction = "horizontal", legend.position = "bottom")
```

### Further questions

* Explore various missing & unknown features in the data
  + What is the cause of missingness in `sector` & `cause`? 
  Are some types of breaches more likely than others to have missing data here?
  + What comprises the `Other` breach cause, as it appeared to be growing?
* I've looked at `sector` & `cause` separately here, but it would be interesting to
look for potential interactions between the two, with respect to breach frequency
and severity (cost).
* Look closer at differences between breaches that involve zero cost/no affected
records and those that do. Can we distinguish any risk factors that might separate
the two?
